---
title: "A Gentle Introduction to Permutation Tests"
layout: post
date: 2015-09-15
comments: false
categories: [Statistics, Data Simulations, R]
---

The usual process when we carry out a hypothesis test is:  
* Define your null hypothesis (H$\sf{_{0}}$) and your alternative hypothesis (H$\sf{_{a}}$);  
* Pick the most likely appropriate test distribution (t, z, binomial, etc.);  
* Compute your test statistic under this distribution;  
* Work out whether we reject the null hypothesis based whether your test statistic exceeds a critical statistic under this distribution.

For example, imagine if we wanted to test whether two types of [example] significantly differed.

The null hypothesis (H0) would be: $\mu\sf{_{1}}$ - $\mu\sf{_{2}}$ = 0, and the alternative hypothesis (Ha) would be: $\mu\sf{_{1}}$ - $\mu\sf{_{2}}$ &#8800; 0

Confidence interval for t-test:
$\bar X\sf{_{1}}$ - $\bar X\sf{_{2}}$ &plusmn; _t_(_n_$\sf{_{1}}$ + _n_$\sf{_{2}}$ - 2), 1 - $\alpha$ / 2 * 

```{r}
data("ToothGrowth")

require(psych)
describe(ToothGrowth$len[ToothGrowth$supp == "OJ"])
describe(ToothGrowth$len[ToothGrowth$supp == "VC"])

sp <- sqrt((30 * 6.61^2 + 30 * 8.27^2)/(30 + 30 - 2))
0 + c(-1, 1) * qt(.975, 28) * sp * (1/30 + 1/30)^.5
20.66 - 16.96

t.test(ToothGrowth$len[ToothGrowth$supp == "OJ"], ToothGrowth$len[ToothGrowth$supp == "VC"], 
       paired = FALSE, var.equal = TRUE)


```

The distribution the critical T and test statistic are calculated under is the distribution of difference between means (as we are trying to see whether difference as extreme is likely under this distribution).

As you can see, a major part of this process is having data that fit the assumptions of your chosen distribution. What if this is not the case? 



```{r, echo=FALSE}
# Generate data
data <- data.frame(group = numeric(length = 100), 
                   null = numeric(length = 100), 
                   alternative = numeric(length = 100))
# Create grouping variable (50 observations in each)
set.seed(567)
data$group <- rep(c(0, 1), c(50, 50))

# Create null (group means do not differ)
data$null <- c(rnorm(50, mean = 10, sd = 2), runif(50, min = 4, max = 16))

# Create alternative (group means differ)
data$alternative[data$group == 0] <- c(rnorm(25, mean = 10, sd = 2), runif(25, min = 4, max = 16))
data$alternative[data$group == 1] <- c(rnorm(25, mean = 10, sd = 2), runif(25, min = 4, max = 16)) + 10

# Create difference between means of groups for null and alternative
null.diff <- mean(data$null[data$group == 1]) - mean(data$null[data$group == 0])
alt.diff <- mean(data$alternative[data$group == 1]) - mean(data$alternative[data$group==0])

# Create a function that randomly reassigns each observation to a different group and then takes the mean difference between these new groups.
one.test <- function(grouping, variable) {
                resampled.group <- sample(grouping)
                mean(variable[resampled.group == 1]) - mean(variable[resampled.group == 0])
            }

# Example of how resampling works:
data$resampled.group <- sample(data$group)
head(data[ , c("group", "resampled.group", "null", "alternative")], n = 10)
mean(data$null[data$resampled.group == 1]) - mean(data$null[data$resampled.group == 0])
null.diff
mean(data$alternative[data$resampled.group == 1]) - mean(data$alternative[data$resampled.group == 0])
alt.diff
data$resampled.group <- NULL

many.truenull <- replicate(1000, one.test(data$group, data$null))
many.falsenull <- replicate(1000, one.test(data$group, data$alternative))

hist(many.truenull)
abline(v=null.diff, lwd=2, col="purple")
mean(abs(many.truenull) > abs(null.diff))

hist(many.falsenull)
abline(v=alt.diff, lwd=2, col="purple")
mean(abs(many.falsenull) > abs(alt.diff))
```



As part of writing this post, I heavily borrowed from the code used in [Thomas Lumley and Ken Rices' presentation](http://faculty.washington.edu/kenrice/sisg/SISG-08-06.pdf) for the Summer Institute in Statistical Genetics, and used code and explanations from [Charlie Geyer's tutorial](http://www.stat.umn.edu/geyer/old/5601/examp/perm.html) from his class at University of Minnesota, Twin Cities, 